Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-28 paper-1/manuscript.tex
Rev sec Results:F7

% TODO rev Fig7
As noted previously, short-term depression alone exerted no deviance detection effect, but its addition did enhance the deviance detection effect established by TA. To understand the basis of this apparent synergy, we contrasted the TA only condition examined in the previous section against the same networks with both TA and STD. On average, adding STD caused a reduction in response magnitude across the board, as we would expect from a depression mechanism (data not shown). However, this reduction was not uniform: Target responses were reduced most in the std sequence, and least in the dev sequence.

To understand the reason for this asymmetry, we first focused on the target response and asked how and why the target response contrast (dev - msc) changed as STD was added to the model. As shown in Figure \ref{fig7}a, the target response was greater in dev than msc in both the full and the ablated model (indicating a positive DDI in both models, cf. also Figure \ref{fig3}). In most neurons, the contrast was greater in the full model, such that the difference (full - ablated) was positive, as indicated by the largely orange plot in the right column.
Following the logic established in the previous section, we may expect a greater target response in dev to be caused by lower $V_{TA}$ relative to msc. Indeed, comparing the contrast (dev - msc) in $V_{TA}$ before target trials across models (Figure \ref{fig7}b), we see that, while the contrast is negative (i.e., lower in dev) in both models, the magnitude of the contrast is greater in the full model. For clarity, the difference is therefore negative, as indicated by the largely blue plot in the right column.
Lastly, we trace this one step further back in Figure \ref{fig7}c: Again as established in the previous section, a greater contrast in $V_{TA}$ may be explained by a similarly enhanced contrast (dev - msc) in the non-target response. We found exactly this: The non-target response was smaller in dev than msc by a greater margin in the full model, as indicated by the largely blue plot in the right column.

This pattern was reflected also in the aggregate across networks and stimuli. Firstly, the contrast (dev - msc) in the target response (Figure \ref{fig7}d, left) was positive in both models, but the full model contrast was greater than the ablated model contrast (t = 2.99, p = 0.00168; one-sided t-test, n = 120). Referring to the previous section, we identify the cause for this in an enhanced contrast in $V_{TA}$, as shown in Figure \ref{fig7}e; the contrast (dev - msc) in the full model was more negative than that in the ablated model (t = -3.49, p = 0.000342). Lastly, returning to Figure \ref{fig7}d (right), we see that the driver of $V_{TA}$, namely the non-target response size, also showed an enhanced contrast (dev - msc) in the full model relative to the ablation (t = -3.19, p = 0.000921). To sum up, the reason for the larger target dev response relative to msc, was the smaller non-target dev response relative to msc; furthermore, since the relationships established in the previous section remained largely unchanged, this effect appeared to be mediated primarily through TA.
% end TODO

\begin{figure*}%
    \centering
    \includegraphics{fig7.pdf}
    \caption{}
    \label{fig7}
\end{figure*}
\textbf{Fig. 7} Adding STD to the model maintains or slighly enhances the effects of TA.
\textbf{a} Median contrast (dev - msc) in the non-target response (top) and in $V_{TA}$ (bottom) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to Figure \ref{fig6}c-d.
\textbf{b} Top: Median contrast (dev - msc) in the non-target response magnitude across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the linear correlation between non-target response contrast and $\Delta V_{TA}$ at the start of target trials across excitatory neurons. Each marker corresponds to one network and target stimulus. Data from the full model with STD and TA, compare Figure \ref{fig5}d. Middle: Non-target response contrast in the full model, plotted against the same contrast in the TA-only model. The red line is the identity function; points above this line represent networks with greater response reduction (dev - msc) in the full model than in the ablated model. Bottom: Subtraction of the two response contrasts.
\textbf{c} Top: Median contrast (dev - msc) in $V_{TA}$ at the start of target trials across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the linear correlation between $\Delta V_{TA}$ and the target response contrast across excitatory neurons. Each marker corresponds to one network and target stimulus. Data from the full model with STD and TA, compare Figure \ref{fig4}e. Middle: $\Delta V_{TA}$ in the full model, plotted against the same contrast in the TA-only model. The red line is the identity function; points above this line represent networks with greater reduction (dev - msc) in $V_{TA}$ in the full model than in the ablated model. Bottom: Subtraction of the two $V_{TA}$ contrasts.

This naturally raises the question of why, with the addition of STD, the non-target dev response was reduced more than other responses. We hypothesized that, due to its relatively short time constant, the direct effect of STD across trials was limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences.

>>>>

As noted previously, STD alone exerted no deviance detection effect, but its addition did enhance the deviance detection effect established by TA. To understand the basis of this apparent synergy, we contrasted the TA only condition examined in the previous section against the same networks evaluated in the full model with both TA and STD. On average, adding STD caused a reduction in response magnitude across the board, as we would expect from a depression mechanism (data not shown). However, this reduction was not uniform: Target responses were reduced most in the std sequence, and least in the dev sequence.

To understand how STD affects deviance detection, we first examined the effects of TA in the full model, following the analysis on the ablated model laid out in the previous section. Starting at the cause of the reduced response in non-target trials, we show in Figure \ref{fig7}a the full model equivalent of Figure \ref{fig6}c and d. The contrasts (dev - msc) in the non-target response magnitude and in the resulting $V_{TA}$ levels at the start of B trials both retained a clear separation between the portion of neurons responding early, which showed increased responses (T = 5.44e+03, p = 1.02e-06; one-sided Wilcoxon, n = 120) and increased TA levels (T = 3.41e+03, p = 9.51e-08), and the neurons responding later, whose response (T = 282, p = 9.19e-19) and TA levels (T = 72, p = 2.49e-17) decreased in the dev sequence. Likewise, the correlation across excitatory neurons between non-target response contrast and the contrast in $V_{TA}$ before B trials remained positive (T = 7.26e+03, p = 9.86e-22) and very strong (median Pearson's $\rho$ = 0.992) This indicates that local adaptation to B mediated by TA remained a key driver of the reduced response to non-target trials.

Inspecting the non-target response reduction in the full model in Figure \ref{fig7}b, we found that it was on average larger than in the ablated model (T = 4.16e+03, p = 0.0175; one-sided Wilcoxon, n = 120), if only by a small amount on the order of 0.03 spikes per neuron. Based on the ablated model analysis in Figure \ref{fig5}, we expected reduced non-target responses to similarly reduce the TA levels at the start of A trials. Indeed, the non-target response contrast remained tightly correlated with the contrast (dev - msc) in $V_{TA}$ at the start of target trials (median Pearson's $\rho$ = 0.989), as shown in Figure \ref{fig7}c. Correspondingly, $V_{TA}$ was reduced in dev, as shown in Figure \ref{fig7}d, and this too by a slighly greater amount in the full model than in the ablated model (T = 4.37e+03, p = 0.0257).

Finally, based on the ablated model analysis in Figure \ref{fig4}, we sought to confirm that lowered $V_{TA}$ before target trials was responsible for the increased target response. As shown in Figure \ref{fig7}e, the contrasts (dev - msc) in $V_{TA}$ at the start of target trials and in response magnitude to A were negatively correlated (T = 30, p = 2.09e-21, median = -0.268; one-sided Wilcoxon, n = 120), though the correlation was weaker in the full model (T = 1.67e+03, p = 1.49e-07). Despite the weaker correlation, however, the full model target response was clearly increased in dev over msc, and this increase was greater than in the ablated model (T = 1.94e+03, p = 0.000723), as shown in Figure \ref{fig7}f. This greater contrast, of course, directly reflects the greater DDI reported in Figure \ref{fig3}.

Taken together, the data indicate that, with STD added to the model, the overall chain of causality -- increased local adaptation to B resulting in reduced B responses and reduced global adaptation, which in turn allows greater target responses -- remained intact with only minor quantitative changes. The primary cause for the greater target response, therefore, appeared to be the smaller non-target response, which was where we noted the first clear shift towards increased deviance detection.

\begin{figure*}%
    \centering
    \includegraphics{fig7.pdf}
    \caption{}
    \label{fig7}
\end{figure*}
\textbf{Fig. 7} Adding STD to the model maintains or slighly enhances the effects of TA.
\textbf{a} Median contrast (dev - msc) in the non-target response (left) and in $V_{TA}$ at the start of B trials (right) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to Figure \ref{fig6}c-d.
\textbf{b} Median contrast (dev - msc) in the non-target response magnitude across networks and stimuli in the full model (horizontal axis) and the ablated model with TA only (vertical axis). The red line is the identity function; points above this line represent networks with greater response reduction (dev - msc) in the full model than in the ablated model. The subtraction of the two quantities in the box plot below shows that the non-target response tends to be reduced more in the full model.
\textbf{c} Non-target response contrast, plotted against its Pearson correlation coefficient with the corresponding contrast in $V_{TA}$ at the start of target trials, in the full model. Compare to Figure \ref{Fig5}d.
\textbf{d} Median contrast (dev - msc) in $V_{TA}$ at the start of target trials across networks and stimuli in the full and ablated models, arranged as in panel b. The subtraction of the two quantities in the box plot below shows that $V_{TA}$ tends to be reduced more in the full model.
\textbf{e} $\Delta V_{TA}$ at the start of target trials, plotted against its Pearson correlation coefficient with the corresponding contrast in the target response magnitude, in the full model. Compare to Figure \ref{Fig4}e.
\textbf{f} Median contrast (dev - msc) in the target response magnitude across networks and stimuli in the full and ablated models, arranged as in panel b. The subtraction of the two quantities in the box plot below shows that the target response tends to be increased more in the full model.

This naturally raises the question of why, with the addition of STD, the non-target dev response was reduced more than the non-target msc responses. We hypothesized that, due to its relatively short time constant, the direct effect of STD across trials was limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences.
